---
title:  "STAT340: Discussion 11: ANOVA"
output: html_document
---

```{r setup, include=FALSE}
# check packages installed
knitr::opts_chunk$set(echo=T)
library(tidyverse)
```


## Exercises

We will practice working with ANOVA models in this discussions.


### 1)

A group of 32 rats were randomly assigned to each of 4 diets labelled A, B, C, and D. The response is the liver weight as a percentage of body weight.

```{r}
liver = read.csv(text="3.42,3.17,3.34,3.65
3.96,3.63,3.72,3.93
3.87,3.38,3.81,3.77
4.19,3.47,3.66,4.18
3.58,3.39,3.55,4.21
3.76,3.41,3.51,3.88",col.names=LETTERS[1:4])

head(liver)
```


a. State your hypotheses and any assumptions you need to make.
b. Construct an ANOVA table manually (using basic R functions) OR using the `aov()` function (recommended to do both ways!)

```{r}
# insert code here

```

c. Interpret the results in the context of the original problem. What is your conclusion?

> ***REPLACE WITH YOUR RESPONSE***

d. **If necessary**, conduct post-hoc analysis to identify which groups are significantly different.
e. Assess your assumptions. Is there any evidence the model results are unreliable? Explain. If you find strong enough evidence of model assumption violations, run a Kruskal-Wallis test as an alternative analysis.


### 2)

We have a dataset of mean improvements in jump height (in inches) from two different college basketball training programs, as well as several other variables. The data is shown below:

```{r}
heights = data.frame(program=rep(c(1, 2), each=20),
                     gender=rep(c('M', 'F'), each=10, times=2),
                     division=rep(c(1, 2), each=5, times=4),
                     height=c(7, 7, 8, 8, 7, 6, 6, 5, 6, 5,
                              5, 5, 4, 5, 4, 3, 3, 4, 3, 3,
                              6, 6, 5, 4, 5, 4, 5, 4, 4, 3,
                              2, 2, 1, 4, 4, 2, 1, 1, 2, 1))

print(heights)
```

Analyze the data to determine which factors have the largest impact on overall height improvement. Repeat parts a-c. and e. from question 1. (Note this time, you're not required to know how to compute manually)

```{r}
# insert code here

```

